Measurement and Probability in Fuzzy Quantum Theories

### 2016-11-29 proposal/prop.tex
almost done

We will begin by reviewing existing work that recasts classical
probability spaces in a resource-aware setting and move to our
proposal which, briefly speaking, aims at recasting quantum
probability and hence quantum measurement to a corresponding
resource-aware setting. In particular, we plan on addressing the
Meyer-Mermin debate~\cite{PhysRevLett.83.3751, Mermin1999,
  BarrettKent2004} on the impact of finite precision measurements on
the relevance of the Kochen-Specker
theorem~\cite{kochenspecker1967,Redhead1987-REDINA,peres1995quantum,Jaeger2007}.
We argue that no matter what the results are their impact will be
strong. Positive results that formulate a computable, with clear
resource bounds, theory of quantum measurement and quantum probability
would be essential for understanding and realizing quantum
computation. Any negative results would redirect research that aims at
realizing quantum computation to other approaches. At an intellectual
level, the results have the potential to solve or clarify several
paradoxes in quantum mechanics, quantum information, computability,
and complexity theory.

@@

%%%%%
\subsection{Proposed Activity IV: Computational Aspects}

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.38]{HoughLineData.pdf}
\end{center}
  \caption{Curves of possible line parameters determined by noisy
    data samples on a straight line in 2D space.}
\label{houghline.fig}
\end{figure}

%%%%%
\subsection{Proposed Activity V: Applications}

A classic evidence-accumulation method known as the Hough
transform~\cite{BallardGenHough1981}, which is very closely related to
Bayesian evidence accumulation
methods~\cite{CrandallEtal2005,CrandallEtal2006}, provides another
example of how our proposal to define physical theories without exact
real numbers could be combined with the measurement process. Simply
described, the Hough transform as an abstraction of Bayesian schemes
starts with a model of the theory being measured, typically a model
that would ordinarily be treated {\it after\/} gathering the data by
applying least squares to choose the optimal parameters.  But in this
method, instead of thinking of, say, the model of a straight line
$\hat{\mathbf{n}}\cdot \mathbf{x} = d$ as an equation for the points
on a line $ \mathbf{x}_{i} = (x_{i},y_{i})$, we think of it as a
time-accumulating {\it set\/} of equations for the distance $d$ of the
nearest point on the line and the direction
$ \hat{\mathbf{n}} = (\cos \theta, \sin \theta)$ from the origin to
that point (the direction perpendicular to the line).  Thus each point
of {\it evidence\/}, $(x_{i},y_{i})$, produces an {\it equation in the
  parameter space\/} that is effectively a plot of all the possible
values of $d(\theta)$ as a function of the possible angular directions
$\theta$ pointing to the family of all possible lines.  A typical
result from a set of errorful (it is impossible to measure real
numbers, we recall) data measurements of points on a hypothetical
straight line is shown in Fig.~\ref{houghline.fig}.

The interesting step for us is now to realize that the {\it parameter
  curves\/} shown in the figure do not need, or indeed, must not, be
themselves infinitely precise real numbers, but they must themselves
incorporate in some way a theory consistent with forbidding absolute
precision.  In practice, then, the sinusoidal lines in Figure
\ref{houghline.fig} all become fuzzy, and the accumulation of these
fuzzy distributions produces peaks of likelihood around the values
$(d_0, \theta_0)$ that are the optimal choices for the line describing
the set of data.  Note that even {\it one\/} data point provides
evidence on a set of hypotheses. The simple analysis we have outlined
here is easily extended to much more complex situations in higher
dimensions (see, e.g.,~\cite{BallardGenHough1981}).  Our point here is
that essentially any physical theory that would ordinarily be
described by a mathematical model using infinite precision numbers is
potentially nonsensical from a computational point of view: computer
scientists know perfectly well, and have been dealing with since the
time of Turing and Shannon, that every single bit of precision in any
number has a cost, and that there is not enough energy in the universe
to compute even one general real number.  Thus the combination of
physical models that, in some intrinsic fashion, contain only
computable numbers, with theories of measurement and evidence in the
fashion of the Hough transform example we have just presented, could
contain the outline of entirely new ways of understanding what
mathematical models of the physical world and its measurement actually
could be.

@@

\paragraph*{Prior Support.} There was no recent NSF support for
Dr. Ortiz or Dr. Hanson. In the past five years, Prof. Sabry received
one NSF grant (award number 1217454) titled \emph{Information Effects}
for the amount of \$275,319.00 with start and end dates of 1 September
2012 -- 31 August 2016.

@@

\paragraph*{Publications.} Several papers were published to report on
the results. On the foundational side, we have the following papers:
(1) Computing with Semirings and Weak Rig Groupoids, ESOP, 2016, (2)
Superstructural Reversible Logic, Workshop on Linearity, 2014, (3)
Theseus: A High Level Language for Reversible
Computing. Work-in-Progress report, RC 2014, (4) Discrete Quantum
Theories, J. Phys. A: Math. Theor. 2014., (5) Geometry of Discrete
Quantum Computing, J. Phys. A: Math. Theor. 2013., (6) Isomorphic
Interpreters from Logically Reversible Abstract Machines. RC 2012. On
the applications side, we have the following papers: (1) Copredication
in Homotopy Type Theory: A homotopical approach for formal semantics
of natural languages. Workshop on Natural Language and Computer
Science, 2016., (2) Contract Monitoring Semantics as Patterns of
Communication. ICFP, 2015., (3) Extensible Effects: An Alternative to
Monad Transformers. Haskell Symposium, 2013., (4) Encoding Secure
Information Flow with Restricted Delegation and Revocation in
Haskell. Workshop on Functional Programming Concepts in
Domain-Specific Languages, 2013. In addition to the publications,
Dr. Sabry and collaborators maintain a public repository with software
related to the project. Dr. Sabry was also invited to give talks at
several venues, most notably the Colloquiumfest, University of
Saskatchewan, Canada, 2014, and the Workshop on Quantum Information
and Foundations of Quantum Mechanics, University of British Columbia,
Vancouver, Canada, 2013

>>>>

We will begin by reviewing existing work that recasts classical
probability spaces in a resource-aware setting and move to our
proposal which, briefly speaking, aims at recasting quantum
probability and hence quantum measurement to a corresponding
resource-aware setting.

In particular, the objectives to be achieved include:
\begin{itemize}
\item A computational framework for quantum measurement where quantum
  systems are described in terms of interactive processes. The
  framework takes care of the resources required (including the ones
  related to the observer) to determine a given property of the
  measured system.
\item Develop a measurement framework based on quantum interval-valued
  probabilities measures.
\item Assess the validity of fundamental theorems of quantum
  mechanics, such as Gleason, Bell, and Kochen-Specker, in a quantum
  formulation stripped away of infinite resources. We will aim to
  provide a definite answer to the Meyer-Mermin
  debate~\cite{PhysRevLett.83.3751, Mermin1999, BarrettKent2004} on
  the impact of finite precision measurements on the relevance of the
  Kochen-Specker
  theorem~\cite{kochenspecker1967,Redhead1987-REDINA,peres1995quantum,Jaeger2007}.
\item Study the computational complexity of quantum theories and
  simulations based on discrete (computable) fields.
\end{itemize}
We argue that no matter what the results are their impact will be
strong. Positive results that formulate a computable, with clear
resource bounds, theory of quantum measurement and quantum probability
would be essential for understanding and realizing quantum
computation. Any negative results would redirect research that aims at
realizing quantum computation to other approaches. At an intellectual
level, the results have the potential to solve or clarify several
paradoxes in quantum mechanics, quantum information, computability,
and complexity theory.

@@

%%%%%
\subsection{Proposed Activity IV: Computational and Simulation Aspects}

@@

% \begin{figure}
% \begin{center}
%   \includegraphics[scale=0.38]{HoughLineData.pdf}
% \end{center}
%   \caption{Curves of possible line parameters determined by noisy
%     data samples on a straight line in 2D space.}
% \label{houghline.fig}
% \end{figure}

% %%%%%
% \subsection{Proposed Activity V: Applications}

% A classic evidence-accumulation method known as the Hough
% transform~\cite{BallardGenHough1981}, which is very closely related to
% Bayesian evidence accumulation
% methods~\cite{CrandallEtal2005,CrandallEtal2006}, provides another
% example of how our proposal to define physical theories without exact
% real numbers could be combined with the measurement process. Simply
% described, the Hough transform as an abstraction of Bayesian schemes
% starts with a model of the theory being measured, typically a model
% that would ordinarily be treated {\it after\/} gathering the data by
% applying least squares to choose the optimal parameters.  But in this
% method, instead of thinking of, say, the model of a straight line
% $\hat{\mathbf{n}}\cdot \mathbf{x} = d$ as an equation for the points
% on a line $ \mathbf{x}_{i} = (x_{i},y_{i})$, we think of it as a
% time-accumulating {\it set\/} of equations for the distance $d$ of the
% nearest point on the line and the direction
% $ \hat{\mathbf{n}} = (\cos \theta, \sin \theta)$ from the origin to
% that point (the direction perpendicular to the line).  Thus each point
% of {\it evidence\/}, $(x_{i},y_{i})$, produces an {\it equation in the
%   parameter space\/} that is effectively a plot of all the possible
% values of $d(\theta)$ as a function of the possible angular directions
% $\theta$ pointing to the family of all possible lines.  A typical
% result from a set of errorful (it is impossible to measure real
% numbers, we recall) data measurements of points on a hypothetical
% straight line is shown in Fig.~\ref{houghline.fig}.

% The interesting step for us is now to realize that the {\it parameter
%   curves\/} shown in the figure do not need, or indeed, must not, be
% themselves infinitely precise real numbers, but they must themselves
% incorporate in some way a theory consistent with forbidding absolute
% precision.  In practice, then, the sinusoidal lines in Figure
% \ref{houghline.fig} all become fuzzy, and the accumulation of these
% fuzzy distributions produces peaks of likelihood around the values
% $(d_0, \theta_0)$ that are the optimal choices for the line describing
% the set of data.  Note that even {\it one\/} data point provides
% evidence on a set of hypotheses. The simple analysis we have outlined
% here is easily extended to much more complex situations in higher
% dimensions (see, e.g.,~\cite{BallardGenHough1981}).  Our point here is
% that essentially any physical theory that would ordinarily be
% described by a mathematical model using infinite precision numbers is
% potentially nonsensical from a computational point of view: computer
% scientists know perfectly well, and have been dealing with since the
% time of Turing and Shannon, that every single bit of precision in any
% number has a cost, and that there is not enough energy in the universe
% to compute even one general real number.  Thus the combination of
% physical models that, in some intrinsic fashion, contain only
% computable numbers, with theories of measurement and evidence in the
% fashion of the Hough transform example we have just presented, could
% contain the outline of entirely new ways of understanding what
% mathematical models of the physical world and its measurement actually
% could be.

@@

\paragraph*{Prior Support.} There was no recent NSF support for
Dr. Ortiz. In the past five years, Dr. Sabry received one NSF grant
(award number 1217454) titled \emph{Information Effects} for the
amount of \$275,319.00 with start and end dates of 1 September 2012 --
31 August 2016.

@@

\paragraph*{Publications.} Several papers were published to report on
the results. On the foundational side, we have the following papers:
\begin{enumerate}
\item Computing with Semirings and Weak Rig Groupoids, ESOP, 2016
\item Superstructural Reversible Logic, Workshop on Linearity, 2014
\item Theseus: A High Level Language for Reversible
  Computing. Work-in-Progress report, RC 2014
\item Discrete Quantum Theories, J. Phys. A: Math. Theor. 2014.
\item Geometry of Discrete Quantum Computing, J. Phys. A:
  Math. Theor. 2013.
\item Isomorphic Interpreters from Logically Reversible Abstract
  Machines. RC 2012. 
\end{enumerate}
On the applications side, we have the following papers: 
\begin{enumerate}
\item Copredication in Homotopy Type Theory: A homotopical approach
  for formal semantics of natural languages. Workshop on Natural
  Language and Computer Science, 2016.
\item Contract Monitoring Semantics as Patterns of
  Communication. ICFP, 2015.
\item Extensible Effects: An Alternative to Monad
  Transformers. Haskell Symposium, 2013.
\item Encoding Secure Information Flow with Restricted Delegation and
  Revocation in Haskell. Workshop on Functional Programming Concepts
  in Domain-Specific Languages, 2013
\end{enumerate}
In addition to the publications, Dr. Sabry and collaborators maintain
a public repository with software related to the project. Dr. Sabry
was also invited to give talks at several venues, most notably:
\begin{itemize}
\item Colloquiumfest, University of Saskatchewan, Canada, 2014
\item Workshop on Quantum Information and Foundations of Quantum
  Mechanics, University of British Columbia, Vancouver, Canada, 2013
\end{itemize}
